---
title: Olympic games medals prediction
author: Daniel Gribel and Joao Paulo Forny
date: \today
output:
  beamer_presentation:
    colortheme: default
    fonttheme: default
    theme: default
  ioslides_presentation: default
subtitle: Predicting host country performance
---

## Model

- Medals history
- Hosting factor
- Performance analysis on recent competitions
- Number of athletes participating
- Population (total and growth rate) -- also check "population pyramids" evolution
- GDP (total, per capita and growth rate)
- Economic freedom
- Sports investment *

## Hosting factor: Total of medals
```{r, echo=FALSE}
source("~/PUC-MSc/datascience/olympics/src/script.R")
bp_totals <- barplot(total_medals, main="Total of medals (1992, 1996, 2000, 2004, 2008, 2012)",
              ylab = "# Medals",
              beside = TRUE,
              col = total_colors,
              border = NA,
              space = c(0.1,0.8),
              ylim = y_lim_total,
              las = 1)

legend("topleft", legend = c("Performance being host", "Performance being visitor"),
       fill = c("#388BFF", "#ADCFFF"),
       bty = "n",
       inset = 0.03)

text(x = bp_totals, y = array(total_medals), label = array(total_medals), pos = 3,
     cex = 0.8, col = "#444444")
```

## Hosting factor: Scores
```{r, echo=FALSE}
colors <- c("#DD4200", "#FF9155")

bp_scores <- barplot(rbind(home_factor, mean_away_factor),
              main="Scores (Host x Away)",
              ylab = "Score",
              beside = TRUE,
              col = colors,
              border = NA,
              space = c(0.1,1),
              ylim = y_lim_score,
              las = 1)

legend("topleft", legend = c("Host score", "Away score (mean)"), fill = colors, bty = "n",
       inset = 0.03)

text(x = bp_scores, y = array(rbind(home_factor, mean_away_factor)), label = array(rbind(home_factor, mean_away_factor)), pos = 3, cex = 0.8, col = "#444444")
```

## Brazil's performance
```{r, echo=FALSE}
medals_bra <- read.table("../dataset/bra-medals.csv", header = T, sep="\t")

cols <- c("#E0D910", "#C8C8C8", "#C7872E")

medals_header <- c("gold", "silver", "bronze")

ylim <- c(0, 1.1*max(medals_bra$gold + medals_bra$silver + medals_bra$bronze))

bp2 <- barplot(t(as.matrix(medals_bra[, medals_header])), main = "Brazil medals distribution (1920 - 2012)", ylab = "# Medals", beside = FALSE, col = cols, border = NA, space = 0.2, names.arg = medals_bra$year, las = 2, ylim = ylim)

legend("topleft", legend = medals_header, fill = cols, bty = "n", inset = 0.03)

text(x = bp2, y = medals_bra$gold + medals_bra$silver + medals_bra$bronze, label = medals_bra$gold + medals_bra$silver + medals_bra$bronze, pos = 3, cex = 0.8, col = "#666666")

```

## Possible achievements

- Predict medals table for the host country
    + How many medals will Brazil win in Rio'2016?
    + What about medals distribution (gold, silver, bronze)?
    + Is it possible to extend such analysis in order to discriminate predictions for specific sports?
- Investigate hosting factor
    + How does hosting factor effectively contributes for performance improvement?
- Explore visualization mechanisms that may bring some insights

## Validation

- Compare prediction results with past olympic games (2012, 2008, 2004, ...), by shifting datasets in time

## Datasets

For medals history:

- The Guardian Sport Datablog (~30K medals entries)
    + http://www.theguardian.com/sport/datablog/2012/jun/25/olympic-medal-winner-list-data

- Sports Reference
    + http://www.sports-reference.com/olympics/

For country indicators:

- The World Bank
    + http://data.worldbank.org
