---
title: "Olympic games medals prediction"
author: "Daniel Gribel and Joao Paulo Forny"
date: \today
output:
  beamer_presentation:
    colortheme: default
    fonttheme: default
    theme: default
subtitle: Predicting host country performance
---

## Model

- Medals history
    - \textcolor{red}{80's problem: USSR, East/West Germany}
- Hosting factor
- GDP (total, per capita and growth rate) and economic freedom
- \textcolor{blue}{Sports investment}
- \textcolor{blue}{Performance analysis on recent competitions}
- \textcolor{red}{Population (total and growth rate)}

## Possible achievements

- Predict medals table for the host country
    + How many medals will Brazil win in Rio'2016?
    + What about medals distribution (gold, silver, bronze)?
    + Is it possible to extend such analysis in order to discriminate predictions for specific sports?
- Investigate hosting factor
    + How does hosting factor effectively contributes for performance improvement?
- Explore visualization mechanisms that may bring some insights

## Validation

- Compare prediction results with past olympic games (2012, 2008, 2004, ...), by shifting datasets in time

## Datasets

### Medals history

- Data Market
    - https://datamarket.com/data/set/24b1/medals-won-at-the-summer-olympic-games

- The Guardian Sport Datablog (~30K medals entries)
    - http://www.theguardian.com/sport/datablog/2012/jun/25/olympic-medal-winner-list-data

- Sports Reference
    - http://www.sports-reference.com/olympics/

### Country indicators

- The World Bank
    - http://data.worldbank.org

## Total of medals (1992 -- 2012)
```{r echo=FALSE}
source("~/PUC-MSc/datascience/olympics/src/script.R")
```
```{r eval=FALSE}
# total of medals for last 6 games 
total_medals <- cbind(total_medals,
    tail(medals[[key_gold]], 6) + 
    tail(medals[[key_silver]], 6) + 
    tail(medals[[key_bronze]], 6))
```
```{r echo=FALSE}
rownames(total_medals) <- paste(years, hosts_acronym)
colnames(total_medals) <- hosts_acronym
total_medals
```

## Total of medals (1992 -- 2012)
```{r, echo=FALSE}
bp_totals <- barplot(total_medals, main="Total of medals (1992, 1996, 2000, 2004, 2008, 2012)",
              ylab = "# Medals",
              beside = TRUE,
              col = total_colors,
              border = NA,
              space = c(0.1,0.8),
              ylim = y_lim_total,
              las = 1)

legend("topleft", legend = c("Performance being host", "Performance being visitor"),
       fill = c("#388BFF", "#ADCFFF"),
       bty = "n",
       inset = 0.03)

text(x = bp_totals, y = array(total_medals), label = array(total_medals), pos = 3,
     cex = 0.8, col = "#444444")
```

## Hosting factor
### Mean away score
```{r eval=FALSE}
# mean away score for last 6 games 
mean_away_factor <- cbind(mean_away_factor,
  mean(
    3*tail(medals[[key_gold]], 6)[-host_pos_tail] +
    2*tail(medals[[key_silver]], 6)[-host_pos_tail] +
    tail(medals[[key_bronze]], 6)[-host_pos_tail]
  ))
```
```{r echo=FALSE}
colnames(mean_away_factor) <- hosts_acronym
mean_away_factor
```

## Hosting factor
### Home score
```{r eval=FALSE}
# home score for last 6 games 
home_factor <- cbind(home_factor,
    3*medals[[key_gold]][host_pos] +
    2*medals[[key_silver]][host_pos] +
    medals[[key_bronze]][host_pos])
```
```{r echo=FALSE}
#colnames(home_factor) <- hosts_acronym
home_factor
```

## Hosting factor
```{r, echo=FALSE}
colors <- c("#DD4200", "#FF9155")

bp_scores <- barplot(rbind(home_factor, mean_away_factor),
              main="Scores (Home x Away)",
              ylab = "Score",
              beside = TRUE,
              col = colors,
              border = NA,
              space = c(0.1,1),
              ylim = y_lim_score,
              las = 1)

legend("topleft", legend = c("Home score", "Away score (mean)"), fill = colors, bty = "n",
       inset = 0.03)

text(x = bp_scores, y = array(rbind(home_factor, mean_away_factor)), label = array(rbind(home_factor, mean_away_factor)), pos = 3, cex = 0.8, col = "#444444")
```

## Hosting factor
```{r eval=FALSE}
home_factor/mean_away_factor
```
```{r echo=FALSE}
round(home_factor/mean_away_factor, 4)
```
```{r}
mean(home_factor/mean_away_factor)
```

## Brazil's performance
```{r, echo=FALSE}
medals_bra <- read.table("../dataset/bra-medals.csv", header = T, sep="\t")

color_medals <- c("#E0D910", "#C8C8C8", "#C7872E")

medals_header <- c("gold", "silver", "bronze")

ylim <- c(0, 1.1*max(medals_bra$gold + medals_bra$silver + medals_bra$bronze))

bp2 <- barplot(t(as.matrix(medals_bra[, medals_header])), main = "Brazil medals distribution (1920 - 2012)", ylab = "# Medals", beside = FALSE, col = color_medals, border = NA, space = 0.2, names.arg = medals_bra$year, las = 2, ylim = ylim)

legend("topleft", legend = medals_header, fill = color_medals, bty = "n", inset = 0.03)

text(x = bp2, y = medals_bra$gold + medals_bra$silver + medals_bra$bronze, label = medals_bra$gold + medals_bra$silver + medals_bra$bronze, pos = 3, cex = 0.8, col = "#666666")

```

## Our first prediction for Brazil medals distribution...
```{r, eval=FALSE}
bra_score <- 
  3*tail(medals$Brazil..Gold, 6) +
  2*tail(medals$Brazil..Silver, 6) +
  tail(medals$Brazil..Bronze, 6)

# ratio: mean between home factor and mean away factor
r <- mean(home_factor/mean_away_factor)

# predicting BRA score
pred_bra_score <- mean(bra_score) * r

# predicting BRA medals distribution
p <- pred_bra_score/3 + pred_bra_score/6 + pred_bra_score/9
pred_bra_gold <- round(bra_gold_perc*p)
pred_bra_silver <- round(bra_silver_perc*p)
pred_bra_bronze <- round(bra_bronze_perc*p)
```

## Our first prediction for Brazil medals distribution...
```{r, echo=FALSE}
ylim_pred <- c(0, 1.2*max(pred_bra))
bp_pred <- barplot(array(pred_bra), col = color_medals, border = NA, ylim = ylim_pred, axes=FALSE)
legend("topleft", legend = medals_header, fill = color_medals, bty = "n", inset = 0.03)
text(x = bp_pred, y = array(pred_bra), label = array(pred_bra), pos = 3, cex = 3.5, col = "#999999")
```

\* HFP: Hosting factor prediction

## GDP
- Analyze correlation between GDP growth and medals growth $\Rightarrow$ fit in a regression model

\begin{center}
$ggf_{c(y)} = \frac{\bar{G}_{c(y-4,y)}}{\bar{G}_{w(y-4,y)}}$
\end{center}

\begin{center}
\center$mgr_{c(y)} = \frac{S_{c(y)}}{S_{c(y-4)}}$
\end{center}

where $\bar{G}_{c(y,y-4)}$ is the average GDP growth on period $[y-4, y[$ for country $c$, $\bar{G}_{w(y,y-4)}$ is the world average GDP growth on period $[y-4, y[$ and $S_{c(y)}$ is the medal score for country $c$ on year $y$.

## Correlation: Medals score x GDP
**Correlation coefficient** of two variables: their covariance divided by the product of their individual standard deviations. It is a normalized measurement of how the two are linearly related.

1: variables are positively linearly related

-1: variables are negatively linearly related

```{r echo=FALSE}
cor_gdp_hosts <- round(subset(cor_gdp, select = hosts), 4)
cor_gdp_bra <- round(cor(country_medals_ratio("Brazil"), ind_factor["Brazil",]), 4)
colnames(cor_gdp_hosts) <- hosts_acronym
colnames(cor_gdp_bra) <- "BRA"
```
```{r}
cor_gdp_hosts
```
```{r}
cor_gdp_bra
```

## Correlation: Medals score x GDP (2012)
```{r, echo=FALSE}
all <- c(hosts, c("Russia", "Germany", "France"))
ggf <- ind_factor[all, match(2012, years)]
mgr <- c()
i <- 1
for(c in all) {
  mgr[i] <- country_medals_ratio(c)[match(2012, years)]
  i <- i+1
}
plot(cbind(mgr,ggf), col=ifelse(rownames(cbind(mgr,ggf))%in% c("China","Great.Britain"), "red", "blue"), pch = 19, cex = 1, ylab = "GDP growth factor", xlab = "Medals growth factor")
text(mgr, ggf, labels=rownames(cbind(mgr,ggf)), cex= 0.7, pos=3)
```

## Projecting GDP: 2016 analysis
```{r, echo=FALSE}
yr <- c(1992:2013)
world <- rowMeans(ind[which(ind$Time %in% yr),][-c(1, which(colnames(ind)=="Brazil"))],na.rm=TRUE)
world <- c(world, c(3.38900, 3.45100))
bra <- tail(ind$Brazil, length(yr))
bra <- c(bra, c(0.14500, -1.02600))
yr <- c(yr, c(2014, 2015))
g_range <- range(0, world, bra)
plot(world, type="o", col="blue", axes=FALSE, ann=TRUE, ylim=g_range, ylab = "GDP", xlab = "Years")
lines(bra, type="o", pch=22, lty=1, col="red")
axis(1, at=1:length(yr), lab=yr)
axis(2, at=-5:10, las=1)
title(main = "World mean GDP x Brazil GDP")
legend("topleft", g_range[2], c("World mean GDP","Brazil GDP"), cex=0.8, col=c("blue","red"), pch=21:22, lty=1:2)

addTrans <- function(color,trans) {
  if (length(color)!=length(trans)&!any(c(length(color),length(trans))==1)) stop("Vector lengths not correct")
  if (length(color)==1 & length(trans)>1) color <- rep(color,length(trans))
  if (length(trans)==1 & length(color)>1) trans <- rep(trans,length(color))

  num2hex <- function(x)
  {
    hex <- unlist(strsplit("0123456789ABCDEF",split=""))
    return(paste(hex[(x-x%%16)/16+1],hex[x%%16+1],sep=""))
  }
  rgb <- rbind(col2rgb(color),trans)
  res <- paste("#",apply(apply(rgb,2,num2hex),2,paste,collapse=""), sep="")
  return(res)
}
lim <- par("usr")
rect(length(yr)-3, lim[3]-1, length(yr), lim[4]+1, border=NA, col=addTrans("red", 50))
box()
```

\tiny{* 2014 and 2015: forecasts}

\tiny{Source: IMF World Economic Outlook (WEO), April 2015}

## Brazil MGR x GGF
```{r, echo=FALSE}
hostData <- groupeddata[!(groupeddata$Country != "Brazil"),]
hostData$Year <- as.numeric.factor(hostData$Year)
# Converting dependent and independent properties (factor to numeric)
hostData$MGR <- as.numeric.factor(hostData$MGR)
hostData$GGF <- as.numeric.factor(hostData$GGF)
plot(hostData$GGF, hostData$MGR, xlab = "Brazil GGF", ylab = "Brazil MGR")
abline(brazil2016LM())
text(hostData$GGF, hostData$MGR, labels=hostData$Year, cex= 0.6, pos=3)
```

## Linear Regression Model for Brazil
```{r, echo=TRUE}
summary(brazil2016LM())
```

## Using predicted MGR to improve 2016 prediction
```{r, echo=TRUE}
intercept = as.numeric(coefficients(brazilLM)[1])
slope = as.numeric(coefficients(brazilLM)[2])

brazil2016MGR = intercept + slope * brazil2016GGF

pred_bra_score * (1 + brazil2016MGR/100)
```

## Sports investment
Brazil:
```{r eval=FALSE}
> sports_investment_bra
  year budget population budget_pp
1 2010   1.50      195.2      7.68
2 2011   2.50      196.9     12.69
3 2012   3.44      198.6     17.32
4 2013   3.38      200.3     16.87
5 2014   3.39      202.7     16.72
6 2015   3.01      204.4     14.72
```

## Sports investment
```{r eval=FALSE}
budget_pp_2016 <- mean(budget_pp[which(year >= 2012)])
> budget_pp_2016
[1] 16.4075
budget_pp_2012 <- mean(budget_pp[which(year < 2012)])
> budget_pp_2012
[1] 10.185
bra_score_2012 <-
     3*tail(medals$Brazil..Gold, 1) +
     2*tail(medals$Brazil..Silver, 1) +
     tail(medals$Brazil..Bronze, 1)
x <- (bra_score_2016*bra_score_2012)/bra_score_2012

> x
[1] 45.10653
```

## Sports investment
Great Britain:
```{r eval=FALSE}
> sports_investment_gbr
  year budget population budget_pp medals_score
1 2000   0.19       58.9      3.22           60
2 2012   0.84       63.7     13.18          140
```
```{r eval=FALSE}
ratio_budget_medals <-
  sports_investment_gbr$budget_pp/sports_investment_gbr$medals_score

> ratio_budget_medals[2]/ratio_budget_medals[1]
[1] 1.754215
```

## Sports investment
```{r eval=FALSE}
sports_inv <- (x-pred_bra_score)/1.754215
> sports_inv
[1] 3.36705

> bra_gold_perc*sports_inv
[1] 0.7488889
> bra_silver_perc*sports_inv
[1] 0.9829167
> bra_bronze_perc*sports_inv
[1] 1.638194
```

Rounding, we have +0 gold medal, +1 silver medal and +2 bronze medals relative to HFP.

## Recent competitions